
---

title: "Linear Regression on mtcars"
author: "Alexander Bruns"
date: "Friday, June 12, 2015"
output:
  html_document: default

---

### Summary

This study reviews the effect of 10 parameters on the gas mileage of 32 cars. The impact of each factor was investigated and modeled. 
It was found that the transmission type has no significant effect on the gas mileage. There are gasoline saving autmomatic automobiles availible as well as gasoline saving manual automobile.
However, these findings only apply to this dataset, because it consists of very fewe observations and besides some exotic cars high hp cars are included; thus they do not represent the average car market.


### Effect of transmission type on mpg based on the *mtcars* dataset

The present study aims to investigate the possible effect of the transmission type in 32 cars represented in the *mtcars* dataset among other variables.

This report is divided into three parts: Exploratory data analysis, Model selection and Model validation.

#### Exploratory Data Analysis

The exploratory data analysis (Appendix A) showed a strong correlation between mpg and weight, displacement and cylinders (R > 0.8). Displacement and cylinders are confounded as usually the displacement increases with cylinder number. Both are correlated with R > 0.9. The correlation between *mpg* and transmission type is about 0.6 which indicates that mpg depends on various factors.

Graph 1 reveals the relation between Weight, Transmission type and Miles per Gallon.  
<center>
```{r echo=FALSE, message=FALSE,warning=FALSE,error=FALSE}
library(ggvis)
mtcars%>%ggvis(x=~wt,y=~mpg)%>%layer_points(fill=~as.factor(am))%>%add_axis("x",title="Weight [wt]")%>%add_axis("y",title="Miles per Gallon [mpg]")%>%set_options(width = 300, height = 300, keep_aspect = TRUE)
```
</center>
#### Model Selection
This section describes the selection of the best model for the question to how extent affects the transmission type the *mpg*. As shown in the previous section, only the variables with the highest correlation coefficient were taken into account. The selection of the best model is done by means of a likelihood-ratio-test.

```{r echo=FALSE, message=FALSE,warning=FALSE,error=FALSE}
fit_all<-lm(mpg~.,data=mtcars)
fit_select<-lm(mpg~wt+am+cyl,data=mtcars)
fit_interact<-lm(mpg~wt+am+cyl+wt*cyl,data=mtcars)
anova(fit_all,fit_select,fit_interact)
```
The result of the ANOVA table shows that the addition of an interaction term increases the accuray. Therefore the linear model *fit_interact* will be used the calculate prediction and confidence intervals.

#### Model Evaluation
The following table shows the summary of the linear model. As it can be seen, the transmission type has no significant effect on mpg. Weight [wt] and cylinders [cyl] are significant regressors. 
In fact, the gas mileage _does not_ depent on the transmission type.
```{r echo=F, message=FALSE,warning=FALSE,error=FALSE}
summary(fit_interact)
```

Residuals (see __Annex C__) are in a good approximation normally distributed. No autocorrelation nor heteroskedasticity could be observed.
However, two car models - the Toyota Corolla and the Fiat 128 - show greater residuals tha expectetd. This might be due to the fact that these small cars were engineered to save gasoline, opposite to the majority of the investigated population.


\newpage 

#### APPENDIX A - Getting and Cleaning the *"mtcars"* Data

__Requiered packages:__
```{r echo=T, message=FALSE,warning=FALSE,error=FALSE}
library(dplyr)
library(lattice)
library(datasets)
library(psych)
```

#### Getting and cleaning the data

```{r echo=T, message=FALSE,warning=FALSE,error=FALSE}
data(mtcars)
```

\newpage 

#### APPENDIX B - Exploratory Data Analysis

#### Inspection for relationship between dependent and independent variables

The dependent variable is *mpg* while the independent variables are weight *wt*, transmission type *am*, cylinder displacement *disp* and horsepower *hp*. 

Firstly, the correlations between all variables are investigated.
```{r echo=T, message=FALSE,warning=FALSE,error=FALSE}
corr.test(mtcars)$r[,1]
```

Then, all variables with a correlation with *mpg* higher than 0.6 are visualized in the following plot in order to examine the type of the correlation.
```{r echo=F, message=FALSE,warning=FALSE,error=FALSE,fig.width=8, fig.height=8}
splom(mtcars[c(1,3,4,6,9)],main="Relation between variables of the dataset")
```

\newpage

#### APPENDIX C - Residuals

The residuals for the model were examined. The following four methods were utilized in order to determine the goodness of fit.

* Residuals vs Fitted

* QQ-Plot to examine normal distribution of residuals

* Standarized residuals

* Residuals vs Leverage

```{r echo=F, message=FALSE,warning=FALSE,error=FALSE,fig.width=8, fig.height=8}
par(mfrow=c(2,2))
plot(fit_interact,pch=16,col='blue')
```

Residuals are in a good approximation normally distributed. No autocorrelation nor heteroskedasticity could be observed.
However, two car models - the Toyota Corolla and the Fiat 128 - show greater residuals tha expectetd. This might be due to the fact that these small cars were engineered to save gasoline, opposite to the majority of the investigated population.
